---
title: "Constructing Experiment Objects with BiocDuckDB"
author:
- name: Patrick Aboyoun
  email: aboyounp@gene.com
  affiliation: Genentech, Inc.
date: "Compiled: `r format(Sys.time(), '%B %d, %Y')`"
package: BiocDuckDB
vignette: >
  %\VignetteIndexEntry{Constructing Experiment Objects with BiocDuckDB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_depth: 3
---

```{r setup, echo=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

# Introduction

The `r Biocpkg("BiocDuckDB")` package enables analysis of genomic datasets too large for memory by storing data in Parquet format and querying via DuckDB's optimized SQL engine. This vignette demonstrates how to work with DuckDB-backed versions of Bioconductor's core experiment containers: *SummarizedExperiment*, *RangedSummarizedExperiment*, and *SingleCellExperiment*.

## The Challenge

Modern genomic experiments routinely generate datasets that exceed available RAM:

- **Single-cell RNA-seq**: 30,000 genes × 1,000,000 cells (sparse)
- **Spatial transcriptomics**: 10,000 genes × 100,000 spots (dense)
- **Multi-omics**: Multiple assays across thousands of samples

Traditional approaches require subsetting data or specialized file formats. BiocDuckDB offers an alternative: keep data on disk while maintaining the familiar Bioconductor API.

## The Solution

BiocDuckDB provides two key functions:

- `writeParquet()`: Serialize Bioconductor objects to Parquet format
- `readParquet()`: Read Parquet files as DuckDB-backed Bioconductor objects

The resulting objects behave identically to their in-memory counterparts, but operations are executed via SQL queries on disk-based data.

```{r load-packages}
library(BiocDuckDB)
library(SummarizedExperiment)
library(SingleCellExperiment)
```


# SummarizedExperiment Workflow

## Basic Round-Trip

The simplest workflow writes an object to Parquet and reads it back:

```{r se-roundtrip}
# Load example RNA-seq data
data(airway, package = "airway")
airway

# Write to Parquet
se_path <- file.path(tempdir(), "airway_se")
writeParquet(airway, se_path)

# Read back as DuckDB-backed
airway_ddb <- readParquet(se_path)
airway_ddb
```

The DuckDB-backed object keeps data on disk, dramatically reducing memory usage.

## Understanding the Storage Layout

`writeParquet()` creates a structured directory with components stored separately:

```{r se-structure}
list.files(se_path, recursive = TRUE)
```

The layout follows the Frictionless Data Package specification:

- `feature_data/`: Gene annotations (rowData)
- `sample_data/`: Sample metadata (colData)
- `assay_data/counts/`: Count matrix in coordinate format
- `datapackage.json`: Metadata describing the schema

## Storage Format: Coordinate Representation

Matrices are stored in **coordinate (COO) format** rather than dense arrays:

```{r coo-format}
# Peek at the assay storage
assay_files <- list.files(file.path(se_path, "assay_data", "counts"), 
                          full.names = TRUE, pattern = "\\.parquet$",
                          recursive = TRUE)
if (length(assay_files) > 0) {
    assay_sample <- arrow::read_parquet(assay_files[1])
    head(assay_sample)
}
```

Each row contains:
- `__feature__`: Gene index (1-based)
- `__sample__`: Sample index (1-based)
- `value`: The actual count

Dimnames are stored as metadata categories, mapping indices to labels. This sparse representation is ideal for single-cell data where most values are zero.

## Operations on DuckDB-Backed Objects

The DuckDB-backed *SummarizedExperiment* supports all standard operations:

```{r se-operations}
# Access components
assay(airway_ddb, "counts")
colData(airway_ddb)
rowData(airway_ddb)[1:5, ]

# Subsetting
airway_sub <- airway_ddb[1:1000, 1:4]
dim(assay(airway_sub, "counts"))

# Statistical operations (pushed to SQL)
colSums(assay(airway_sub, "counts"))
rowMeans(assay(airway_ddb, "counts"))[1:10]
```

## Memory Comparison

```{r memory-comparison}
# In-memory object
print(object.size(airway), units = "MB")

# DuckDB-backed object (only metadata in RAM)
print(object.size(airway_ddb), units = "MB")

# Reduction factor
as.numeric(object.size(airway) / object.size(airway_ddb))
```


# RangedSummarizedExperiment with Genomic Coordinates

## Preserving Genomic Annotations

The airway dataset includes genomic coordinates for each gene:

```{r rse-coords}
# Check object class
class(airway)
class(airway_ddb)

# Genes are annotated with exon coordinates
rowRanges(airway)[[1]]
```

`writeParquet()` automatically converts *GRangesList* objects to Parquet LIST[] columns, preserving all genomic information.

## Genomic Operations

The DuckDB-backed version supports standard *GRanges* operations:

```{r rse-operations}
# Access row ranges (returns DuckDBGRangesList)
rr_ddb <- rowRanges(airway_ddb)
rr_ddb

# Query genomic features
elementNROWS(rr_ddb)[1:10]  # Exons per gene
seqinfo(airway_ddb)         # Chromosome information

# Subsetting preserves structure
airway_chr1 <- airway_ddb[1:100, ]
airway_chr1
```


# SingleCellExperiment for Large-Scale Single-Cell Data

## Motivation

Single-cell experiments are where BiocDuckDB shines. A typical 10x Genomics experiment with 1 million cells:

- **In memory**: ~30 GB (sparse matrix + metadata)
- **Parquet storage**: ~3 GB (10x compression)
- **DuckDB-backed RAM**: ~50 MB (metadata only)

## Configuration for Large Sparse Matrices

Configure block processing before writing large datasets:

```{r sce-config}
library(DelayedArray)
library(scRNAseq)

# Process data in 1 GB blocks
setAutoBlockSize(size = 1e9)
setAutoBlockShape("scale")
```

## Round-Trip Example

```{r sce-roundtrip}
# Load brain single-cell data (subset for demonstration)
sce <- ZeiselBrainData()
sce <- sce[1:2000, 1:1000]
sce

# Write to Parquet (handles sparse matrix efficiently)
sce_path <- file.path(tempdir(), "zeisel_sce")
writeParquet(sce, sce_path)

# Read back as DuckDB-backed
sce_ddb <- readParquet(sce_path)
sce_ddb
```

## Storage Efficiency

```{r sce-efficiency}
# In-memory size
sce_mem <- object.size(sce) / 1024^2
cat(sprintf("In-memory: %.2f MB\n", sce_mem))

# On-disk size
sce_disk <- sum(file.info(
    list.files(sce_path, full.names = TRUE, recursive = TRUE)
)$size) / 1024^2
cat(sprintf("Parquet: %.2f MB\n", sce_disk))

# DuckDB-backed RAM
sce_ddb_mem <- object.size(sce_ddb) / 1024^2
cat(sprintf("DuckDB-backed: %.2f MB\n\n", sce_ddb_mem))

# Compression and memory reduction
cat(sprintf("Disk compression: %.1fx\n", sce_mem / sce_disk))
cat(sprintf("Memory reduction: %.1fx\n", sce_mem / sce_ddb_mem))
```

## Single-Cell Analysis Operations

All standard operations work on DuckDB-backed objects:

```{r sce-operations}
# Access metadata
colData(sce_ddb)[1:5, 1:5]

# QC filtering (SQL-optimized)
high_quality <- sce_ddb[, colData(sce_ddb)$total_counts > 5000]
cat(sprintf("Retained %d / %d cells\n", ncol(high_quality), ncol(sce_ddb)))

# Feature selection
gene_means <- rowMeans(assay(sce_ddb, "counts"))
hvg <- sce_ddb[gene_means > 1, ]
cat(sprintf("Selected %d / %d genes\n", nrow(hvg), nrow(sce_ddb)))

# Statistical summaries (pushed to SQL)
total_counts <- colSums(assay(sce_ddb, "counts"))
summary(total_counts)
```


# Recommended Analysis Workflow

## The Filter-Realize-Analyze Pattern

For datasets too large for memory, follow this pattern:

1. **Write** raw data to Parquet
2. **Load** as DuckDB-backed
3. **Filter** using SQL-optimized operations
4. **Realize** the filtered subset to memory
5. **Analyze** using standard Bioconductor tools
6. **Write** results back to Parquet

```{r workflow-example, eval=FALSE}
library(scuttle)
library(scran)
library(scater)

# 1-2. Write and load large dataset
writeParquet(large_sce, "data/raw_counts")
sce_ddb <- readParquet("data/raw_counts")

# 3. Filter efficiently on disk (SQL-optimized)
qc_metrics <- colData(sce_ddb)
keep_cells <- qc_metrics$sum > 1000 & 
              qc_metrics$detected > 500 &
              qc_metrics$subsets_Mito_percent < 5
sce_filtered <- sce_ddb[, keep_cells]

# Keep only genes expressed in ≥10 cells
gene_detection <- rowSums(assay(sce_filtered, "counts") > 0)
sce_filtered <- sce_filtered[gene_detection >= 10, ]

# 4. Realize filtered subset (now manageable in memory)
sce_clean <- as(sce_filtered, "SingleCellExperiment")

# 5. Standard analysis pipeline
sce_clean <- logNormCounts(sce_clean)
dec <- modelGeneVar(sce_clean)
hvg <- getTopHVGs(dec, n = 2000)
sce_clean <- runPCA(sce_clean[hvg, ], ncomponents = 50)
sce_clean <- runUMAP(sce_clean, dimred = "PCA")

# 6. Save processed results
writeParquet(sce_clean, "data/processed_counts")
```

This pattern enables analysis of datasets 10-100x larger than available RAM.

## When to Use DuckDB-Backed Objects

**Ideal for**:

- Initial data exploration and QC
- Filtering and subsetting large datasets
- Computing summary statistics
- Storage and distribution
- Datasets that don't fit in memory

**Not ideal for**:

- Iterative algorithms with random access patterns
- Dense matrix operations (better to realize subset first)
- Small datasets (<1 GB) that fit comfortably in memory

## Performance Optimization

```{r optimization-demo}
# Efficient: Filter first, then realize
high_expr <- rowMeans(assay(sce_ddb, "counts")) > 1
sce_subset <- sce_ddb[high_expr, ]
sce_mem <- as(sce_subset, "SingleCellExperiment")

cat(sprintf("Filtered: %d → %d genes\n", nrow(sce_ddb), nrow(sce_subset)))
cat(sprintf("Memory footprint: %.2f MB\n", 
            object.size(sce_mem) / 1024^2))
```


# Multi-Omics with MultiAssayExperiment

For multi-omics experiments, store each modality separately:

```{r mae-workflow, eval=FALSE}
library(MultiAssayExperiment)

# Write each modality to its own directory
mae_path <- file.path(tempdir(), "multiomics")
writeParquet(rna_sce, file.path(mae_path, "rna"))
writeParquet(protein_se, file.path(mae_path, "protein"))
writeParquet(atac_sce, file.path(mae_path, "atac"))

# Read back as DuckDB-backed experiments
experiments <- list(
    rna = readParquet(file.path(mae_path, "rna")),
    protein = readParquet(file.path(mae_path, "protein")),
    atac = readParquet(file.path(mae_path, "atac"))
)

# Construct MultiAssayExperiment
mae <- MultiAssayExperiment(
    experiments = experiments,
    colData = shared_sample_data
)
```

Each modality stays on disk until accessed, enabling memory-efficient multi-omics analysis.


# Advanced Features

## Custom Index Column Names

Specify meaningful column names for better interpretability:

```{r custom-cols, eval=FALSE}
writeParquet(sce, output_path, 
             indexcols = c("gene_id", "cell_barcode"))
```

## Partition Pruning with Dimension Tables

For very large datasets, dimension tables enable efficient filtering:

```{r dimtables, eval=FALSE}
# writeParquet automatically creates dimension tables
# These enable partition pruning in DuckDB queries
counts <- assay(large_sce, "counts")
dimtbls <- createDimTables(counts)  # Encodes data structure

writeParquet(large_sce, output_path)  # Uses dimtbls internally
```

## Block Processing

Configure block processing for matrices too large to realize at once:

```{r block-config, eval=FALSE}
# Process in 1 GB chunks
setAutoBlockSize(1e9)
setAutoBlockShape("scale")

writeParquet(very_large_sce, output_path)

# Operations use block processing automatically
col_totals <- colSums(assay(sce_ddb, "counts"))
```


# Format Comparison: Parquet vs RDS

Parquet offers several advantages over traditional RDS serialization:

```{r format-comparison}
# Compare storage formats
rds_file <- tempfile(fileext = ".rds")
saveRDS(airway, rds_file)

comparison <- data.frame(
    Format = c("RDS", "Parquet"),
    Size_MB = c(
        file.info(rds_file)$size / 1024^2,
        sum(file.info(list.files(se_path, full.names = TRUE, 
                                  recursive = TRUE))$size) / 1024^2
    )
)
comparison$Compression <- with(comparison, 
                                Size_MB[1] / Size_MB)
comparison
```

**Parquet advantages**:

- **Columnar compression**: 3-10x smaller for sparse data
- **Cross-language**: Readable by Python, Julia, JavaScript
- **Selective loading**: Access specific columns without loading entire file
- **Schema metadata**: Self-describing format
- **Query optimization**: DuckDB can push filters to Parquet metadata

**RDS advantages**:

- **Simplicity**: Single function call
- **Exact serialization**: Preserves all R object attributes
- **Random access**: Better for dense data with complex access patterns


# Best Practices

## Data Provenance

Store metadata about data processing:

```{r provenance}
metadata(airway)$creation_date <- Sys.time()
metadata(airway)$genome_build <- "GRCh38"
metadata(airway)$processing_notes <- "Filtered for genes with mean count > 10"
metadata(airway)$package_version <- packageVersion("BiocDuckDB")

# Metadata is preserved in Parquet files
writeParquet(airway, file.path(tempdir(), "annotated_airway"))
```

## Recommended Directory Structure

Organize large projects systematically:

```
project/
├── data/
│   ├── raw/              # Original data (writeParquet)
│   ├── qc/               # After QC filtering
│   ├── normalized/       # Normalized data
│   └── processed/        # Final analysis results
├── metadata/
│   └── sample_info.csv   # Sample metadata
└── results/
    └── analysis.Rmd      # Analysis notebooks
```

## Version Control

Track data versions using directory names:

```{r versioning, eval=FALSE}
# Include date or version in path
writeParquet(sce, sprintf("data/experiment_v%s_%s", 
                          version, Sys.Date()))
```


# Session Information

```{r sessioninfo}
sessionInfo()
```
