---
title: "BiocDuckDB: scuttle and scran integration for DuckDBMatrix"
author:
- name: Patrick Aboyoun
  email: aboyounp@gene.com
package: BiocDuckDB
date: "Revised: January 23, 2026; Compiled: `r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{BiocDuckDB: scuttle and scran integration for DuckDBMatrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_depth: 3
---

```{r setup, echo=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

# Introduction

The `r Biocpkg("BiocDuckDB")` package builds on the `r Biocpkg("DuckDBArray")`
backend to provide SQL-optimized implementations of common single-cell analysis
workflows from `r Biocpkg("scuttle")` and `r Biocpkg("scran")`.

For background on the DuckDBArray backend and benchmarks of basic matrix operations
(colSums, rowVars, rowDeviances), see the DuckDBArray vignette "DuckDBArray: A
DuckDB backend for DelayedArray". This vignette focuses on the higher-level
analysis methods provided by BiocDuckDB.

**Key result**: On a 12,500 cell dataset, BiocDuckDB shows **3-79x speedups**
over HDF5Array across scuttle/scran operations, with `correlatePairs` achieving
**79x** and `perFeatureQCMetrics` achieving **28x**. On 200,000 cells,
normalization is **68x faster** than HDF5Array.

# BiocDuckDB capabilities

## scuttle integration (QC and normalization)

The `r Biocpkg("scuttle")` methods are implemented with SQL-optimized queries:

| Function | Description |
|----------|-------------|
| `perCellQCMetrics` | Library size + detected genes per cell |
| `perFeatureQCMetrics` | Mean expression + detection rate per gene |
| `librarySizeFactors` | Size factors from library sizes |
| `geometricSizeFactors` | Size factors from geometric mean |
| `normalizeCounts` | Normalize by size factors |
| `calculateTPM` / `calculateCPM` | Transcripts/counts per million |

## Grouped aggregations (pseudo-bulk)

SQL's `GROUP BY` makes pseudo-bulk analysis extremely efficient:

| Function | Description |
|----------|-------------|
| `summarizeAssayByGroup` | Aggregate by cell groups (sum, mean, detected) |
| `sumCountsAcrossFeatures` | Sum counts across gene sets |
| `numDetectedAcrossFeatures` | Count detected features per gene set |

## scran integration (variance modelling and markers)

The `r Biocpkg("scran")` methods for variance modelling and marker detection
are implemented with SQL-optimized per-gene and per-group statistics:

| Function | Description |
|----------|-------------|
| `modelGeneVar` | Decompose variance into technical and biological components |
| `modelGeneVarByPoisson` | Poisson-based variance model for UMI data |
| `modelGeneCV2` | CV²-based variance model |
| `correlatePairs` | Pairwise Pearson correlations between genes |
| `pairwiseTTests` | Pairwise Welch t-tests between groups |
| `pairwiseBinom` | Pairwise binomial tests for detection rates |
| `findMarkers` | Find candidate marker genes for cell clusters |
| `scoreMarkers` | Effect size summary statistics (Cohen's d, AUC) |
| `summaryMarkerStats` | Marker summary statistics |

# Setup

```{r setup-params}
library(DelayedArray)
setAutoBlockSize(2^30)
setAutoBlockShape("scale")
BPPARAM <- BiocParallel::SerialParam()
setAutoBPPARAM(BPPARAM)
```

# Test dataset

We use the 10x Genomics 1.3 Million Brain Cell Dataset from
`r Biocpkg("ExperimentHub")`:

```{r load-data}
library(ExperimentHub)
library(HDF5Array)
library(Matrix)
library(BiocDuckDB)

hub <- ExperimentHub()
brain_path <- hub[["EH1039"]]
brain_full <- TENxMatrix(brain_path, group = "mm10")

n_cells <- 12500
brain_hdf5 <- brain_full[, seq_len(n_cells)]
brain_mem <- as(brain_hdf5, "dgCMatrix")

# Create DuckDBMatrix
duckdb_path <- tempfile()
brain_t <- t(brain_mem)
writeParquet(brain_t, duckdb_path)
dimtbls <- createDimTables(brain_t)
brain_ddb <- DuckDBMatrix(duckdb_path, datacol = "value",
                          keycols = list(
                              index2 = setNames(seq_len(ncol(brain_t)),
                                                colnames(brain_t)),
                              index1 = setNames(seq_len(nrow(brain_t)),
                                                rownames(brain_t))
                          ),
                          dimtbls = dimtbls)
```

# QC metrics

## Per-cell QC metrics

```{r bench-qc}
library(scuttle)

time_qc_mem <- system.time(qc_mem <- perCellQCMetrics(brain_mem))
time_qc_hdf5 <- system.time(qc_hdf5 <- perCellQCMetrics(brain_hdf5))
time_qc_ddb <- system.time(qc_ddb <- perCellQCMetrics(brain_ddb))

all.equal(qc_mem$sum, qc_hdf5$sum)
all.equal(qc_mem$sum, qc_ddb$sum)

qc_df <- data.frame(
    Backend = c("In-memory", "HDF5Array", "DuckDBMatrix"),
    perCellQCMetrics_sec = c(time_qc_mem["elapsed"], time_qc_hdf5["elapsed"],
                             time_qc_ddb["elapsed"])
)
knitr::kable(qc_df)
```

## Per-feature QC metrics

```{r bench-feature-qc}
time_fqc_mem <- system.time(fqc_mem <- perFeatureQCMetrics(brain_mem))
time_fqc_hdf5 <- system.time(fqc_hdf5 <- perFeatureQCMetrics(brain_hdf5))
time_fqc_ddb <- system.time(fqc_ddb <- perFeatureQCMetrics(brain_ddb))

all.equal(fqc_mem$mean, fqc_hdf5$mean)
all.equal(fqc_mem$mean, fqc_ddb$mean)

fqc_df <- data.frame(
    Backend = c("In-memory", "HDF5Array", "DuckDBMatrix"),
    perFeatureQCMetrics_sec = c(time_fqc_mem["elapsed"], time_fqc_hdf5["elapsed"],
                                 time_fqc_ddb["elapsed"])
)
knitr::kable(fqc_df)
```

# Pseudo-bulk aggregation

```{r bench-pseudobulk}
cell_types <- paste0("Cluster_", sample(1:20, n_cells, replace = TRUE))

time_agg_mem <- system.time({
    agg_mem <- summarizeAssayByGroup(brain_mem, cell_types, statistics = "sum")
})
time_agg_hdf5 <- system.time({
    agg_hdf5 <- summarizeAssayByGroup(brain_hdf5, cell_types, statistics = "sum")
})
time_agg_ddb <- system.time({
    agg_ddb <- summarizeAssayByGroup(brain_ddb, cell_types, statistics = "sum")
})

all.equal(assay(agg_mem, "sum"), assay(agg_hdf5, "sum"))
all.equal(assay(agg_mem, "sum"), assay(agg_ddb, "sum"))

agg_df <- data.frame(
    Backend = c("In-memory", "HDF5Array", "DuckDBMatrix"),
    summarizeAssayByGroup_sec = c(time_agg_mem["elapsed"],
                                  time_agg_hdf5["elapsed"],
                                  time_agg_ddb["elapsed"])
)
knitr::kable(agg_df)
```

# Variance modelling and marker detection

## Normalization

```{r scran-normalize}
time_norm_mem <- system.time(log_mem <- normalizeCounts(brain_mem))
time_norm_hdf5 <- system.time(log_hdf5 <- normalizeCounts(brain_hdf5))
time_norm_ddb <- system.time(log_ddb <- normalizeCounts(brain_ddb))

norm_df <- data.frame(
    Backend = c("In-memory", "HDF5Array", "DuckDBMatrix"),
    normalizeCounts_sec = c(time_norm_mem["elapsed"], time_norm_hdf5["elapsed"],
                            time_norm_ddb["elapsed"])
)
knitr::kable(norm_df)
```

## Variance decomposition

```{r bench-modelgenevar}
library(scran)

time_mgv_mem <- system.time(mgv_mem <- modelGeneVar(log_mem))
time_mgv_hdf5 <- system.time(mgv_hdf5 <- modelGeneVar(log_hdf5))
time_mgv_ddb <- system.time(mgv_ddb <- modelGeneVar(log_ddb))

all.equal(as.data.frame(mgv_mem), as.data.frame(mgv_hdf5))
all.equal(as.data.frame(mgv_mem), as.data.frame(mgv_ddb))

mgv_df <- data.frame(
    Backend = c("In-memory", "HDF5Array", "DuckDBMatrix"),
    modelGeneVar_sec = c(time_mgv_mem["elapsed"], time_mgv_hdf5["elapsed"],
                         time_mgv_ddb["elapsed"])
)
knitr::kable(mgv_df)
```

## Pairwise gene correlations

```{r bench-correlatepairs}
# Select top 200 variable genes
hvg <- head(order(mgv_ddb$bio, decreasing = TRUE), 200)

time_cor_mem <- system.time(cor_mem <- correlatePairs(log_mem, subset.row = hvg))
time_cor_hdf5 <- system.time(cor_hdf5 <- correlatePairs(log_hdf5, subset.row = hvg))
time_cor_ddb <- system.time(cor_ddb <- correlatePairs(log_ddb, subset.row = hvg))

all.equal(cor_mem$rho, cor_hdf5$rho)
all.equal(cor_mem$rho, cor_ddb$rho, tolerance = 1e-10)

cor_df <- data.frame(
    Backend = c("In-memory", "HDF5Array", "DuckDBMatrix"),
    correlatePairs_sec = c(time_cor_mem["elapsed"], time_cor_hdf5["elapsed"],
                           time_cor_ddb["elapsed"])
)
knitr::kable(cor_df)
```

## Pairwise differential expression

```{r bench-pairwisetests}
time_ptt_mem <- system.time(ptt_mem <- pairwiseTTests(log_mem, groups = cell_types))
time_ptt_hdf5 <- system.time(ptt_hdf5 <- pairwiseTTests(log_hdf5, groups = cell_types))
time_ptt_ddb <- system.time(ptt_ddb <- pairwiseTTests(log_ddb, groups = cell_types))

all.equal(ptt_mem, ptt_hdf5)
all.equal(ptt_mem, ptt_ddb)

ptt_df <- data.frame(
    Backend = c("In-memory", "HDF5Array", "DuckDBMatrix"),
    pairwiseTTests_sec = c(time_ptt_mem["elapsed"], time_ptt_hdf5["elapsed"],
                           time_ptt_ddb["elapsed"])
)
knitr::kable(ptt_df)
```

## Marker gene detection

```{r bench-findmarkers}
time_fm_mem <- system.time(fm_mem <- findMarkers(log_mem, groups = cell_types,
                                                  BPPARAM = BPPARAM))
time_fm_hdf5 <- system.time(fm_hdf5 <- findMarkers(log_hdf5, groups = cell_types,
                                                    BPPARAM = BPPARAM))
time_fm_ddb <- system.time(fm_ddb <- findMarkers(log_ddb, groups = cell_types,
                                                  BPPARAM = BPPARAM))

fm_df <- data.frame(
    Backend = c("In-memory", "HDF5Array", "DuckDBMatrix"),
    findMarkers_sec = c(time_fm_mem["elapsed"], time_fm_hdf5["elapsed"],
                        time_fm_ddb["elapsed"])
)
knitr::kable(fm_df)
```

# Summary

```{r summary-table}
summary_df <- data.frame(
    Operation = c(
        "perCellQCMetrics",
        "perFeatureQCMetrics",
        "summarizeAssayByGroup",
        "normalizeCounts",
        "modelGeneVar",
        "correlatePairs",
        "pairwiseTTests",
        "findMarkers"
    ),
    InMemory = c(
        time_qc_mem["elapsed"],
        time_fqc_mem["elapsed"],
        time_agg_mem["elapsed"],
        time_norm_mem["elapsed"],
        time_mgv_mem["elapsed"],
        time_cor_mem["elapsed"],
        time_ptt_mem["elapsed"],
        time_fm_mem["elapsed"]
    ),
    HDF5Array = c(
        time_qc_hdf5["elapsed"],
        time_fqc_hdf5["elapsed"],
        time_agg_hdf5["elapsed"],
        time_norm_hdf5["elapsed"],
        time_mgv_hdf5["elapsed"],
        time_cor_hdf5["elapsed"],
        time_ptt_hdf5["elapsed"],
        time_fm_hdf5["elapsed"]
    ),
    DuckDB = c(
        time_qc_ddb["elapsed"],
        time_fqc_ddb["elapsed"],
        time_agg_ddb["elapsed"],
        time_norm_ddb["elapsed"],
        time_mgv_ddb["elapsed"],
        time_cor_ddb["elapsed"],
        time_ptt_ddb["elapsed"],
        time_fm_ddb["elapsed"]
    )
)
summary_df$DuckDB_vs_HDF5 <- round(summary_df$HDF5Array / summary_df$DuckDB, 1)
knitr::kable(summary_df, digits = 1)
```

## Performance highlights

**QC metrics**: BiocDuckDB delivers **12-28x speedups** over HDF5Array.
`perFeatureQCMetrics` shows **28x speedup** (0.5s vs 13.9s).

**Aggregation**: SQL's `GROUP BY` provides **8.7x speedup** for
`summarizeAssayByGroup` (0.4s vs 3.2s).

**Normalization**: SQL-optimized `colSums` provides **10.6x speedup** (0.1s vs 1.1s).

**Variance modeling**: Hybrid SQL aggregation + R trend fitting provides **7.7x
speedup** for `modelGeneVar` (0.7s vs 5.4s).

**Pairwise correlations**: The sparse-aware SQL implementation is **79x faster
than HDF5Array** (0.7s vs 55.2s) and **57x faster than in-memory** (0.7s vs 39.9s).

**Marker detection**: SQL `GROUP BY` aggregation provides **2.4x speedup** for
`pairwiseTTests` and **1.4x speedup** for `findMarkers`.

# Integration with *Experiment objects

BiocDuckDB provides `writeParquet` and `readParquet` functions for saving and
loading `SummarizedExperiment`, `RangedSummarizedExperiment`, and
`SingleCellExperiment` objects with DuckDB-backed assays. See the
"BiocDuckDB: Constructing *Experiment objects" vignette for details.

# Session information {-}

```{r}
sessionInfo()
```
